---
title: "p8105_hw2_jkl2195"
author: "Jessie Li"
date: "2023-10-01"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(readxl)
```

Import datasets
```{r}
df_pols_month = read.csv("data/pols-month.csv",)
df_unemployment = read.csv("data/unemployment.csv")
df_snp = read.csv("data/snp.csv")

```

## Problem 1
Clean up pols-month.csv
```{r}
df_pols_month = df_pols_month |>
  separate(mon, c("year", "month", "day"), "-", convert = TRUE)

df_pols_month = df_pols_month |>
  mutate(
    month = month.abb[as.numeric(pull(df_pols_month, "month"))],
    president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")
  )|>
  select(-prez_dem, -prez_gop, -day)# Convert month number to month name
```

Clean up snp.csv
```{r}
df_snp = df_snp |>
  separate(date, c("month", "day", "year"), "/", convert = TRUE)  |>
  mutate(year = case_when((year < 24) ~ year + 2000,
                             TRUE ~ year + 1900))

df_snp = df_snp |>
  mutate(
    month = month.abb[as.numeric(pull(df_snp, "month"))]
  ) |>
  select(-day) |>
  arrange(year, month)
```

Clean up unemployment.csv
```{r}
df_unemployment = df_unemployment |>
  rename(year = "Year") |>
  pivot_longer(Jan:Dec, names_to = "month", values_to = "unemployment_percent")
```

Merging snp and unemployment into pols
```{r}
df_pols_snp_merged = df_pols_month |> 
  left_join(df_snp) |>
  left_join(df_unemployment)
```
**Description**  
The resulting data set contains the party of the president and the number of governors, senator, and representative in each party. There are total of `r nrow(df_pols_snp_merged)` of entries. The data range between `r min(pull(df_pols_snp_merged, year))` to `r max(pull(df_pols_snp_merged, year))` and is organized by month. The data also indicate the closing values of the S&P stock, a value that evaluates the economy at the point, and the unemployment rate in percentage.

## Problem 2
```{r}
df_mr_trash = read_excel("data/202309 Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel", range = "A2:N586") |>
  janitor::clean_names("snake") |>
  mutate(
    wheel = "Mr. Trash Wheel",
    year = as.numeric(year)
  ) |>
  select(-dumpster)

df_prof_trash = read_excel("data/202309 Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel", range = "A2:M108") |>
  janitor::clean_names("snake") |>
  mutate(
    wheel = "Professor Trash Wheel",
    year = as.numeric(year)
  ) |> 
  select(-dumpster)

df_gwynnda_trash = read_excel("data/202309 Trash Wheel Collection Data.xlsx", sheet = "Gwynnda Trash Wheel", range = "A2:L157") |>
  janitor::clean_names("snake") |>
  mutate(
    wheel = "Gwynnda Trash Wheel",
    year = as.numeric(year)
  ) |>
  select(-dumpster)


df_trash_merged = df_mr_trash |>
  bind_rows(df_prof_trash, df_gwynnda_trash) |>
  mutate(
    homes_powered = if_else(homes_powered == 0, weight_tons / 30 * 500, homes_powered)
  )

df_trash_merged
```
**Description**  
This dataset records the weight and volume of trash picked up by Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda Trash Wheel. Trash is seperated to There are a total of `r nrow(df_trash_merged)`. This dataset does not contain the number of dumpster registered but keeps all the categories of the waste. Some key categories includes the weight of plastic, number of plastic bottles, and the weight of polystyrene. Gwynnda Trash Wheel have missing values in categories plastic bottles, wrapper, home powered and sports ball. Professor Trash Wheel having missing values in sports balls. Mr Trash Wheel has some missing values in home powered variable. Based on the note given by original dataset variable `homes_powered` denotes each ton of trash equates to on average 500 kilowatts of electricity. The home_powered for missing values are calculated with the assumption an average household will use 30 kilowatts per day. The total weight of trash collected by Professor Trash Wheel was `r pull(df_prof_trash, weight_tons) |> sum()` tons. The total number of cigarette butts collected by Gwynnda in July of 2021 is `r filter(df_gwynnda_trash, year == 2021) |> pull(cigarette_butts) |> sum() |> format(scientific = FALSE)`.

# Problem 3
**MCI Baseline**
```{r}
df_mci_baseline = read_csv("data/MCI_baseline.csv", skip = 1, na = c(".", "NA")) |>
  janitor::clean_names("snake") |>
  mutate(
    sex = case_match(
      sex,
      0 ~ "female", 
      1 ~ "male"
      ),
    apoe4 = case_match(
      apoe4,
      0 ~ "non carrier",
      1 ~ "carrier"
    ),
    current_age = as.numeric(current_age),
    age_at_onset = as.numeric(age_at_onset)
  ) |>
  #Remove any participants who do not meet the stated inclusion criteria (i.e. no MCI at baseline)
  filter(current_age <= age_at_onset | is.na(age_at_onset))

```
**Description**  
The MCI baseline data set contains the demographics of individuals who are likely to get Mild Cognitive Impairment of Alzheimer’s disease. The Individuals are also recorded if they are a carrier of APOE4, a gene that significantly associated with a higher risk of developing Alzheimer's disease. Demographics include ag e at the beginning of the study, sex, and years of education. Sex and the appearence of APOE4 are recoded to the corresponding index. Individuals that are already in Mild Cognitive Impairment before the study are dropped. In total of `r nrow(df_mci_baseline)` participants that are qualified, `r pull(df_mci_baseline, age_at_onset) |> na.omit() |> length()` individuals develop MCI? The average baseline age is `r pull(df_mci_baseline, current_age) |> mean()` The proportion of women in the study are APOE4 carriers is `r filter(df_mci_baseline, apoe4 == "carrier" & sex == "female") |> nrow() / nrow(df_mci_baseline)`.

**MCI Amyloid**
```{r}
# Both Na and NA is treated as missing value.
df_mci_amyloid = read_csv("data/mci_amyloid.csv", na = c("NA", "Na"), skip = 1) |>
  janitor::clean_names("snake") |>
  rename(
    id = study_id
  )
```
**Description**  
The MCI amyloid contains the study id of the individuals, time since the beginning of the study to conduct Amyloid 42/40 testing, and biomarker values at corresponding time. 

**Compare demographic and biomarker datasets**  
The individuals with the following study id are only in baseline/demographic dataset:

```{r}
df_mci_baseline |> 
  select(id) |> 
  filter(!(id %in% pull(df_mci_amyloid, id))) |> 
  pull() |> 
  print()
```


The individuals with the following study id are only in amyloid/biomarker dataset:
```{r}
df_mci_amyloid |> 
  select(id) |> 
  filter(!(id %in% pull(df_mci_baseline, id))) |> 
  pull() |> 
  print()
```

**Combine demographic and biomarker datasets and save them inside data folder**  
```{r}
df_mci_merged = inner_join(df_mci_baseline, df_mci_amyloid)
write.csv("data/mci merged.csv", row.names = FALSE)
```
**Description**  
The combined data consist of `r nrow(df_mci_merged)` individuals combined from `r nrow(df_mci_baseline)` from the baseline data set and `r nrow(df_mci_amyloid)` individuals from the amyloid dataset.

